---
title: "Population Projection Matrices"
runtime: shiny
output:
  flexdashboard::flex_dashboard:
    navbar:
      - { title: "reproducible.co.nz", href: "https://reproducible.co.nz", align: right }
    orientation: columns
    vertical_layout: fill
    theme: spacelab
    source_code: embed 
---
<style>

body {
  padding-top: 70px;
}


.navbar-brand {
font-family: "Candara";
  font-weight: bold;
  font-size: 24px;
}


</style>

```{r}
# invisible(if(!("pacman" %in% list.files(.libPaths()))) install.packages("pacman"))

# load via pacman required packages
# pacman::p_load("sf", "flexdashboard", "shiny", "tidyverse", "DT")

library(flexdashboard)
library(shiny)
library(tidyverse)
library(shinyMatrix)
library(popbio)
library(diagram)
library(DT)


# turn off messages globally


```


Matrix
=====================================



Column {.sidebar data-width=250}
-------------------------------------

**Matrix input**
```{r}
numericInput("stages", label = "Number of ages / stages:",  "2")

```

<br/>

**Life history diagram: features**
```{r}
sliderInput("bubble.size", label = "Bubble size", min = 0, 
        max = 0.05, value = 0.05, ticks = FALSE)
```

<br/>

**Finite rate of growth**
*(lambda)*

```{r}
renderPrint({
  
  req(input$stages)
  req(input$matrix1)
  
  cat(round(lambda(input$matrix1), 4))
})
```
<br/>

**Time units required for population doubling**

*(only if population is growing)*

```{r}
renderPrint({
  
  req(input$stages)
  req(input$matrix1)
  
  my.lambda <- lambda(input$matrix1)
  
  if(my.lambda > 1) {
  cat(round(log(2, lambda(input$matrix1)), 2))
  } else {
  cat(NA)
  }
})
```
<br/>



```{r}
# Create placeholder for the download button
uiOutput("download_matrix_UI")

output$download_matrix_UI <- renderUI( {
  downloadButton("down_matrix_Btn", "Download matrix", style = "width:100%;")
})

output$down_matrix_Btn <- downloadHandler(
  filename = function() {'Name your matrix.csv'},
  content = function(file) {
    write.csv(input$matrix1, file, row.names = TRUE)
  }
)
```


<br/>

```{r}
# Create placeholder for the download button
plotInput <- reactive({
  options(warn = -1)
  
  req(input$stages)
  req(input$matrix1)
  req(input$bubble.size)
  
  my.matrix <- suppressMessages(input$matrix1)
  
  validate(need(dim(my.matrix)[1] == input$stages, "Updating..."))
  
  rownames(my.matrix) <- paste("stage", 1:input$stages)
  colnames(my.matrix) <- paste("stage", 1:input$stages)
  
  plotmat(my.matrix,
          input$stages,
          box.size = input$bubble.size,
          curve = 0.5)
  
})


uiOutput("download_diagram_UI")

output$download_diagram_UI <- renderUI({
  downloadButton("down_diagram_Btn", "Download diagram", style = "width:100%;")
})

output$down_diagram_Btn <- downloadHandler(
  filename = "Name this diagram.png",
  content = function(file) {
    png(file, width = 800, height = 700, units = "px")
    plotInput()
    dev.off()
  }
)
```



Column {data-width=460}
-------------------------------------

### Matrix of vital rates (survival and fecundity)

* Columns represent the stage the species transitions **from**
* Rows represent the stage the species transitions **to**

<br/>

```{r}

renderUI({
  
  req(input$stages)
  
    ir <- input$stages
    ic <- input$stages
    validate(
      need(isTruthy(ir),label = "Number of ages /stages is required"),
    )
    
    my.matrix <- matrix(data=0,
                               nrow=ir,
                               ncol=ic)
    colnames(my.matrix) <- paste("Stage", 1: input$stages)
    rownames(my.matrix) <- paste("Stage", 1: input$stages)
    
    matrixInput("matrix1",
                value = my.matrix,
                rows = list(
                  n=ir, names = TRUE,
                  editableNames = TRUE),
                cols = list (
                  n=ic, names = TRUE,
                  editableNames = TRUE),
                class = "numeric")
  })

```





Column {data-width=700}
-------------------------------------
### Life history diagram

```{r}



my.diagram <- reactive({
  
  options(warn=-1)

 req(input$stages)
 req(input$matrix1)
 req(input$bubble.size)


  
 my.matrix <- suppressMessages(input$matrix1)

 
validate(
  need(
    dim(my.matrix)[1]== input$stages, "Updating...")
)

 rownames(my.matrix) <- paste("stage", 1: input$stages)
 colnames(my.matrix) <- paste("stage", 1: input$stages)
 

 plotmat(my.matrix, input$stages,  box.size = input$bubble.size,  curve = 0.5)

 
 
 
})



renderPlot({
  
  my.diagram()
  
})






```

Population structure
=====================================

Column {.sidebar data-width=250}
-------------------------------------

**Stable age/stage structure**:  the expected stage/age structure if the vital rates are sustained over the long term.

<br/>

**Reproductive value**: the comparative reproductive value of an individual in each stage/age assuming the vital rates are sustained over the long term. The units are in juvenile equivalents (where a juvenile represents the earliest possible stage).



Column 
-------------------------------------

### Stable age structure

```{r}

renderPrint({
  
  req(input$stages)
  req(input$matrix1)
  
  sa <- stable.stage(input$matrix1) %>% round(3) * 100
  
  sa.df <- data.frame(stage = 1: input$stages , percentage = sa)
  
  print(sa.df, row.names = FALSE)
})
```

### Reproductive value

```{r}

renderPrint({
  
  req(input$matrix1)
  
  ra <- reproductive.value(input$matrix1) %>% round(3)
  
  ra.df <- data.frame(stage = 1: input$stages , reproductive.value = ra)
  
  print(ra.df, row.names = FALSE)
})
```


Diagnostics
=====================================




Column {.sidebar data-width=250}
-------------------------------------

**Elasticity** = results of a small
proportional perturbation

<br/>

**Sensitivity** = results of a small
incremental perturbation


*Warning: Users of this dashboard should be familiar with the consequences of pre- and post breeding monitoring, and birth pulse vs birth flow dynamics on matrix structure. Failure to account for such differences will lead to misinterpretation.*



Column 
-------------------------------------




### Elasticity

```{r}

renderPrint({
  
  req(input$matrix1)
  
  elasticity(input$matrix1) %>% round(3)
})
```

### Sensitivity

```{r}

renderPrint({
  
  req(input$matrix1)
  
  sensitivity(input$matrix1) %>% round(3)
})
```



Population projections
=====================================

Column {.sidebar data-width=250}
-------------------------------------



**Edit starting seeds**
(for matrix projection)

```{r}
renderUI({
  
  req(input$stages)
  
  ir <- input$stages

  validate(
    need(isTruthy(ir),label = "Matrix is required"),
  )
  
  my.matrix <- matrix(data=0, nrow=ir)
  rownames(my.matrix) <- paste("Stage", 1: input$stages)
  
  matrixInput("vector",
              value = my.matrix,
              rows = list(
                n=ir, names = TRUE,
                editableNames = TRUE),
              cols = list(
                n=1, names = FALSE,
                editableNames = TRUE),
              class = "numeric")
})
```

<br/>

```{r}
sliderInput("show.years", HTML("<b>Maximum years to show</b>"),
    min = 0, max = 20, value = 20
  )
```



<br/>
<br/>

*Warning: the graph is showing a projection, what would happen if the vital rates were stable over time. Projections are useful for setting ball-park expectations when carrying capacity is not an issue, but should not be used for general forecasting.*

<br/>

```{r}
uiOutput("download_graph_UI")

output$download_graph_UI <- renderUI({
  downloadButton("down_graph_Btn", "Download graph", style = "width:100%;")
})

output$down_graph_Btn <- downloadHandler(
  filename = "Name this projection.png",
    content = function(file) {
      png(file=file, width = 900, height = 600)
      plot(my.plot())
      dev.off()
  }
)
```


Column 
-------------------------------------

```{r}
# make projection

my.plot <-   reactive({
  
  req(input$matrix1)
  req(input$vector)
  
  
  validate(
  need(ncol(input$matrix1) == length(input$vector), "Updating...")
)

  my.start <- input$vector %>% as.vector()
  my.proj <- pop.projection(input$matrix1, my.start)
  
  # projection
  p <- pop.projection(input$matrix1, my.start)
  
  # change data structure
  projection <- p$stage.vectors %>% as.data.frame()
  projection <- t(projection)
  colnames(projection) <- paste("Stage", 1:dim(projection)[2])
  projection <- projection %>% as.data.frame()
  projection$Year <- 0:(nrow(projection) - 1)
  projection <- projection %>% relocate(Year, .before = 1)
  
  # pivot longer
  projection.longer <- projection %>%
    pivot_longer(cols = colnames(projection)[2]:colnames(projection)[ncol(projection)],
                 names_to = "Stage",
                 values_to = "N")
  
  # filter based on input$show.years
  projection.longer <- projection.longer %>% filter(Year <= input$show.years)
  
  # remove "Stage" string from variable
  projection.longer$Stage <- str_replace_all(projection.longer$Stage , "Stage", "")
  projection.longer$Stage <- as.numeric(as.character(projection.longer$Stage)) %>% as.factor()

  # graph
  ggplot() +
    theme_bw() +
    geom_col(data = projection.longer,
              aes(x = Year, y = N, fill = Stage)) +
    xlab("\nYears since establishment") +
    ylab("Projected population\n")+
    theme(axis.text = element_text(size =14))+
    theme(axis.title = element_text(size =20))+
    theme(legend.text = element_text(size =14))+
    theme(legend.title = element_text(size =14))+
    scale_x_continuous(breaks = 0:max(projection.longer$Year))
  
})
  


#activate <- eventReactive(input$submit, {my.plot()})

renderPlot({
  
  my.plot()
  
}, height = 700)

```  



Videos: dashboard
=====================================

Column {.sidebar data-width=250}
-------------------------------------

Video help can be found on the righthand side. 

If you are running this dashboard locally and not seeing anything try opening the dashboard in your browser (by clicking "Open in Browser" present in the top left of this screen).

Alternatively, click the links below: 

 * [Tour: Introduction](https://www.loom.com/share/4f0b50882e9c4e9988fc16bd97a9237c?sid=5df8d81b-905c-43a9-8c35-7c2b07e83d33)
 * [Tour: Matrix tab](https://www.loom.com/share/fbba17a736584bafbd2b9cd6cc2784e5?sid=f00ad79e-12ee-4612-bbc7-713743aaec0a)
 * [1: Intro to PPMs](https://www.loom.com/share/e64d7f6b7636407c80deffd305c9a22b?sid=49ef4cc5-8f71-4e6f-a2d5-57674ec95708)
 * [2: Analysing PPMs](https://www.loom.com/share/4bca6673af5e4f9e8f0ef6f832e1eef4?sid=9eb1f84e-5be6-4f36-98f5-01696b59035f)
 * [3: Birthing](https://www.loom.com/share/24d60f0e5be148cb9713158b954b501b?sid=02e1c797-cc8c-4b1d-8ede-2e08cbc5559c)
 * [4: Finding data](https://www.loom.com/share/c911e57fdcf64c8a8a340ea5e56bdf5e?sid=5e495545-dc44-4eaf-ba7c-983f4299ce27)

Column {.tabset .tabset-fade}
-------------------------------------

### Tour: Introduction

<div style="position: relative; padding-bottom: 56.25%; height: 0;"><iframe src="https://www.loom.com/embed/4f0b50882e9c4e9988fc16bd97a9237c?sid=5ce6847b-61cc-4f9c-89ad-e9cbe3c42cd5" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 75%; height: 75%;"></iframe></div>


### Tour: Matrix tab

<div style="position: relative; padding-bottom: 56.25%; height: 0;"><iframe src="https://www.loom.com/embed/fbba17a736584bafbd2b9cd6cc2784e5?sid=eafcd485-96ed-41e0-9f7e-cca495948f09" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 75%; height: 75%;"></iframe></div>

### 1:Intro to PPMs

<div style="position: relative; padding-bottom: 56.25%; height: 0;"><iframe src="https://www.loom.com/embed/e64d7f6b7636407c80deffd305c9a22b?sid=e5a4cd13-6a0b-4aa5-929f-e25b2e529144" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 75%; height: 75%;"></iframe></div>


### 2: Analysing PPMs

<div style="position: relative; padding-bottom: 56.25%; height: 0;"><iframe src="https://www.loom.com/embed/4bca6673af5e4f9e8f0ef6f832e1eef4?sid=ca629469-2205-48b7-b83d-30c54b2c08c4" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 75%; height: 75%;"></iframe></div>

### 3: Birthing

<div style="position: relative; padding-bottom: 56.25%; height: 0;"><iframe src="https://www.loom.com/embed/24d60f0e5be148cb9713158b954b501b?sid=6c1192c9-4301-4304-b870-ebed33c9bc8d" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 75%; height: 75%;"></iframe></div>

### 4: Finding data

<div style="position: relative; padding-bottom: 56.25%; height: 0;"><iframe src="https://www.loom.com/embed/c911e57fdcf64c8a8a340ea5e56bdf5e?sid=cc9f3ff9-14ac-4484-b757-0811bea19acb" frameborder="0" webkitallowfullscreen mozallowfullscreen allowfullscreen style="position: absolute; top: 0; left: 0; width: 75%; height: 75%;"></iframe></div>



Reference 
=====================================


Column {.tabset .tabset-fade}
-------------------------------------

```{r}
tags$iframe(style="height:1000px; width:100%; scrolling=yes",
src="Kendall et al 2019.pdf")
```


